---
title: "Exploring the BRFSS data"
author: "Michael Park"
date: "2026 February 12"
output:
  pdf_document: default
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

```{r load-data}
load("brfss2013.RData")
```

### Dimensions of the dataset
```{r dimesions}
dim(brfss2013)
```
There are 491,775 cases and 330 vectors.


* * *

## Part 1: Data

The Behavioral Risk Factor Surveillance System (BRFSS) is a collaborative effort between 
the CDC and all 50 states, District of Columbia, and other U.S. territories to gather data
on individual health related issues through telephone survey. The goal of BRFSS is to gather
data by state concerning preventative health practices and risk behaviors linked to chronic
diseases, injuries, and preventable infectious diseases. Telephone surveys are done through
landline and cellular line.

For landline surveys, a random adult was chosen per household. For cellular telephone,
individuals who have private residence or college housing were chosen. Both forms of survey
deployments involved the non-institutionalized population. Optional questions were included
beginning in 2013 involving childhood health for those 17 or younger.  
\

#### Scope of Inference: Generalizability and Causality  

Only those who were non-institutionalized were able to take the survey,
therefore the data reflects a more narrow population than otherwise. The population 
that does not have access to landlines or cellphones are not represented. Finally, the
491,775 surveyed may only reflect those that are willing and/or have the time to take
the survey.

BRFSS resembles a cross-sectional survey, therefore only inferences of
correlation may be made, not causality. Because of the way the survey is deployed,
there may exist recall bias, e.g. asking for health information regarding the past
30 days. There may also exist proxy bias, as adults answer for children. Finally,
self-reporting bias can exist as information cannot be verified.


* * *

## Part 2: Research questions

**Research quesion 1:** Are there differences in the severity of diabetes across ethnic groups?
Are there, also, differences between sex? This question is of interest, because it may reveal 
groups that are at risk of advanced diabetes.

<!-- vectors: _race, diabete3, insulin. -->

**Research quesion 2:** Is there an association between the days reported of mental health
not good and the number of hours slept? The question is of interest because if an association
is found, further investigation into causality could improve treatments for sleep or mental health.
<!-- vectors: sleeptim1, menthlth -->

**Research quesion 3:** In 2013, is there disparity in the income bracket of $75,000 or more between 
sex? This question is important to ask because it may reveal efficacy of equality policies at all 
levels concerning the workplace.
<!--vectors: X_state, sex, income2 -->


* * *

## Part 3: Exploratory data analysis


**Research quesion 1:**

```{r}
d_prop_summary <- brfss2013 %>% 
  filter(!is.na(X_race), diabete3 == "Yes", 
  insulin %in% c("Yes", "No")) %>%
  group_by(X_race) %>%
  summarize(
    n = n(), 
    p_hat = mean(insulin == "Yes"), 
    se = sqrt(p_hat * (1 - p_hat) / n),
    lower = p_hat - 1.96 * se,
    upper = p_hat + 1.96 * se
  ) %>%
  arrange(desc(p_hat))
d_prop_summary

```
The above summary table shows proportion of insulin takers with diabetes, in descending order,  grouped by ethnicity.
\  
 

```{r}
ggplot(d_prop_summary, 
       aes(x = p_hat, y = X_race)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = lower, xmax = upper),
                width = 0.2) +
  scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 28)) +
  labs(
    x = "Proportion taking insulin",
    y = "Race/Ethnicity",
  )
```

The above graph displays a point plot with 95% confidence intervals based on proportion of insulin takers grouped by ethnicity.  
\

```{r}
d_prop_summary_sex <- brfss2013 %>% 
  filter(!is.na(X_race), diabete3 == "Yes", 
  insulin %in% c("Yes", "No")) %>%
  group_by(X_race, sex) %>%
  summarize(
    n = n(), 
    p_hat = mean(insulin == "Yes"), 
    se = sqrt(p_hat * (1 - p_hat) / n),
    lower = p_hat - 1.96 * se,
    upper = p_hat + 1.96 * se
  )
d_prop_summary_sex
```
The above summary table shows porportion of insulin takers with diabetes grouped by ethnicity and sex.


```{r}
ggplot(d_prop_summary_sex, 
       aes(y = X_race, x = p_hat, color = sex, group = sex)) +
  geom_point(position = position_dodge(width = 0.6), 
             size = 2) +
  geom_errorbar(aes(xmin = lower, xmax = upper),
                position = position_dodge(width = 0.6),
                width = 0.2) +
  scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 28)) +
  labs(
    y = "Race/Ethnicity",
    x = "Proportion taking insulin",
    color = "Sex"
  )

```

The above graph displays proportion of insulin takers grouped by race and sex along with confidence intervals.
\

**Research quesion 2:**

```{r}
mhealth_sleep <- brfss2013 %>%
  filter(!is.na(menthlth),
         !is.na(sleptim1),
         menthlth <= 30,
         sleptim1 > 0) %>%
  group_by(menthlth) %>%
  summarize(
    mean_sleep = mean(sleptim1),
    n = n(),
  )
print(mhealth_sleep, n = Inf)
```



```{r}

ggplot(mhealth_sleep, aes(x = menthlth, y = mean_sleep)) +
  geom_point() +
  geom_line() +
  geom_smooth(se = FALSE) +
  labs(x = "Days mental health not good", y = "Mean sleep hours")

```



**Research quesion 3:**

```{r}
income_sex <- brfss2013 %>%
  filter(!is.na(sex), 
         !is.na(income2),
        ) %>%
  group_by(X_state, sex) %>%
  summarize(phat = mean(income2 == "$75,000 or more"), .groups = "drop")
print(income_sex, n = Inf)
  
```

The above summary table shows grouping by state and sex and the respective proportion that earns an income of $75,000 or more.

```{r, fig.height = 10, fig.width = 8}
ggplot(income_sex, 
       aes(x = phat, y = reorder(X_state, phat), color = sex, group = sex)) +
  geom_line(aes(group = X_state), color = "grey70") + # adds horizontal "track"
  geom_point(position = position_dodge(width = 0.6),
             size = 2) +
labs(x = "Proportion of $75,000 or more income earners",
     y = "State",
     color = "Sex")
```

The above point plot shows grouping by state and sex with the proportion of $75,000 or greater earners arranged in descending order.