---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

```{r load-data}
load("brfss2013.RData")
dim(brfss2013)
```



* * *

## Part 1: Data

The Behavioral Risk Factor Surveillance System (BRFSS) is a collaborative effort between 
the CDC and all 50 states, District of Columbia, and other U.S. territories to gather data
on individual health related issues through telephone survey. The goal of BRFSS is to gather
data by state concerning preventative health practices and risk behaviors linked to chronic
diseases, injuries, and preventable infectious diseases. Telephone surveys are done through
landline and cellular line.

For landline surveys, a random adult was chosen per household. For cellular telephone,
individuals who have private residence or college housing were chosen. Both forms of survey
deployments involved the non-institutionalized population. Optional questions were included
beginning in 2013 involving childhood health for those 17 or younger.  
\

#### Scope of Inference: Generalizability and Causality  

Only those who were non-institutionalized were able to take the survey,
therefore the data reflects a more narrow population than otherwise. The population 
that does not have access to landlines or cellphones are not represented. Finally, the
491,775 surveyed may only reflect those that are willing and/or have the time to take
the survey.

BRFSS resembles a cross-sectional survey, therefore only inferences of
correlation may be made, not causality. Because of the way the survey is deployed,
there may exist recall bias, e.g. asking for health information regarding the past
30 days. There may also exist proxy bias, as adults answer for children. Finally,
self-reporting bias can exist as information cannot be verified.






* * *

## Part 2: Research questions

**Research quesion 1:** Are there differences in the severity of diabetes across ethnic groups?
This is question of interest, because it may reveal groups that are at risk of advanced diabetes.
Core variables: _race, diabete3, insulin.

**Research quesion 2:** Are there accessibility differences in frequency of doctor visits across
ethnic groups? This may reveal disparity in access to health care per ethnicity.
Core variables: drvisits, _race

**Research quesion 3:** Is there an association between the amount of time sleeping and mental health?
Core variables: sleptim1, menthlth


* * *

## Part 3: Exploratory data analysis


**Research quesion 1:**

```{r}
diabetes2013 <- brfss2013 %>% filter(!is.na(X_race), 
  diabete3 == "Yes", insulin %in% c("Yes", "No"))
diabetes2013 %>% group_by(X_race) %>% summarize(prop_insulin = 
  sum(insulin == "Yes") / n()) 

```
Summary table showing proportion of insulin taken grouped by ethnicity with diabetes.

```{r}
diabetes2013 %>%
  filter(insulin %in% c("Yes", "No")) %>%
  ggplot(aes(x = X_race, fill = insulin)) +
  geom_bar(position = position_fill(reverse = TRUE)) +
  labs(x = "Ethnicity", y = "Proportion", fill = "Insulin") +
  coord_flip()
```

Proportion of insulin taken vs not taken grouped by ethnicity in a bar graph.


**Research quesion 2:**

```{r}

```



**Research quesion 3:**

```{r}

```

